---
title: "GLMM Algorithms"
author: "Mukai Wang"
date: "6/15/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro

I notice that the `glmer` function in `lme4` package has two algorithm setting parameters. 

* `nAGQ` declares the number of points per axis for evaluating the adaptive Gauss-Hermite approximation to the log likelihood. When `nAGQ` is 0 or 1, the log likelihood will be approximated by [Laplace approximation](https://stats.stackexchange.com/questions/499269/different-estimates-for-mixed-effects-logistic-regression-and-pwrssupdate-error). Note that `nAGQ` being zero means penalized iteratively reweighted least square. When `nAGQ` is larger than zero, nonlinear optimization will be applied.

* There are two nonlinear optimizing algorithms available: `bobyqa` and `Nelder_Mead`. 

The default setting is `nAGQ` being 1 (Nelder Mead algorithm on log likelihood calculated by Laplace Approximation).


# Analysis on All Taxa

I run all five possible combinations of likelihood approximation and optimization algorithm for all 1653 pairs of taxa.

```{r Preparation, echo=FALSE, message=FALSE}
library(lme4)
library(ggplot2)
library(dplyr)
library(DiffRatio)
rm(list=ls())
model_folder <- '/home/wangmk/UM/Research/MDAWG/Differential_Ratio/DiffRatio/RMD/CAARS_Model_Summary'
data_folder <- '/home/wangmk/UM/Research/MDAWG/Differential_Ratio/DiffRatio/RMD/CAARS_data'
```

```{r readdata}
counts_table <- read.csv(file.path(data_folder, 'filtered_count.csv'))
metadata <- read.csv(file.path(data_folder, 'filtered_metadata.csv'))
```

We have already filtered the data to get rid of samples with very shallow sequencing depth and taxa which has structural zeros. There are `r nrow(counts_table)` individuals and `r ncol(counts_table)` taxa. `r sum(metadata$asthma)` samples come from patients with asthma. Before I run GLMM I apply Wilcoxon test and T test on all pairs of taxa first.

```{r Wilcoxon_T_test}
taxa_names <- colnames(counts_table)
taxa_pairs <- combn(taxa_names, 2) %>% t()
pvals_result <- as.data.frame(taxa_pairs)
colnames(pvals_result) <- c("Taxa1", "Taxa2")

#########  Wilcoxon and T test ##############
# raise all values by one
shifted_counts_table <- counts_table + 1
asthma_info <- metadata$asthma
sample_id <- metadata$SAMPLE_ID

for (j in 1:nrow(pvals_result)){
  # generate a subset dataframe
  t1_name <- pvals_result$Taxa1[j]
  t2_name <- pvals_result$Taxa2[j]
  subset_data <- cbind(shifted_counts_table[, t1_name],
                       shifted_counts_table[, t2_name],
                       asthma_info) %>% as.data.frame()
  colnames(subset_data) <- c(t1_name, t2_name, "Asthma")
  subset_data$logratio <- log(subset_data[, t1_name])-log(subset_data[, t2_name])

  # wilcoxon test
  wilx_test <- wilcox.test(logratio~Asthma, data=subset_data)
  pvals_result$wilx[j] <- wilx_test$p.value

  # t test
  t_test <- t.test(logratio~Asthma, data=subset_data)
  pvals_result$ttest[j] <- t_test$p.value
}
```

Then apply GLMM.

```{r GLMM, warning=FALSE}
if (file.exists(file.path(model_folder, 'Pvals.csv'))){
  pvals_result <- read.csv(file.path(model_folder, 'Pvals.csv'))
} else{
  pvals_result$GLMM_PIRLS <- 0
  pvals_result$GLMM_LA_NM <- 0
  pvals_result$GLMM_LA_BOBYQA <- 0
  pvals_result$GLMM_GQ_NM <- 0
  pvals_result$GLMM_GQ_BOBYQA <- 0 
  
  for (j in 1:nrow(pvals_result)){
  t1 <- pvals_result$Taxa1[j]
  t2 <- pvals_result$Taxa2[j]

  # penalized iteratively reweighted least squares
  glmm_PIRLS_model <- dfr(count_table=counts_table, sample_info=metadata,
                     covar=c("asthma"), tpair=c(t1, t2), reff="SAMPLE_ID", taxa_are_rows = FALSE,
                     nAGQ = 0L) |> summary()

  pvals_result$GLMM_PIRLS[j] <- glmm_PIRLS_model$coefficients[[8]]

  # nelder mead, LA
  glmm_NM_LA <- dfr(count_table=counts_table, sample_info=metadata,
                     covar=c("asthma"), tpair=c(t1, t2), reff="SAMPLE_ID", taxa_are_rows = FALSE,
                     nAGQ = 1L, optimizer = "Nelder_Mead") |> summary()
  pvals_result$GLMM_LA_NM[j] <-glmm_NM_LA$coefficients[[8]]


  # nelder mead, nagq=10
  glmm_NM_GQ <- dfr(count_table=counts_table, sample_info=metadata,
                      covar=c("asthma"), tpair=c(t1, t2), reff="SAMPLE_ID", taxa_are_rows = FALSE,
                      nAGQ = 10L, optimizer = "Nelder_Mead") |> summary()
  pvals_result$GLMM_GQ_NM[j] <- glmm_NM_GQ$coefficients[[8]]


  # bobyqa, LA
  glmm_bobyqa_LA <- dfr(count_table=counts_table, sample_info=metadata,
                         covar=c("asthma"), tpair=c(t1, t2), reff="SAMPLE_ID", taxa_are_rows = FALSE,
                         nAGQ = 1L, optimizer = "bobyqa") |> summary()
  pvals_result$GLMM_LA_BOBYQA[j] <- glmm_bobyqa_LA$coefficients[[8]]

  # bobyqa, nagq=10
  glmm_bobyqa_GQ <- dfr(count_table=counts_table, sample_info=metadata,
                          covar=c("asthma"), tpair=c(t1, t2), reff="SAMPLE_ID", taxa_are_rows = FALSE,
                          nAGQ = 10L, optimizer = "bobyqa") |> summary()
  pvals_result$GLMM_GQ_BOBYQA[j] <- glmm_bobyqa_GQ$coefficients[[8]]
  }
  
  write.csv(pvals_result, file.path(model_folder, 'Pvals.csv'), row.names=FALSE)
}
```

One observation is that different algorithm settings for GLMM sometimes generate surprising different results, as I show below.

```{r discrepancy}
GLMM_results <- pvals_result[, seq(5, 9)]
GLMM_results$variance <- apply(GLMM_results, 1, var)

# rank the variance of p values between different algorithms
tpair_pvalvar_rank <- length(GLMM_results$variance) + 1 - rank(GLMM_results$variance)
rowids <- which(tpair_pvalvar_rank %in% seq(1, 5))
selected_pairs <- pvals_result[rowids, ]
knitr::kable(selected_pairs, digits=3)
```

Let's look at two specific pairs: `r selected_pairs$Taxa1[4]`/ `r selected_pairs$Taxa2[4]` and `r selected_pairs$Taxa1[5]`/`r selected_pairs$Taxa2[5]`.


# Closer look


##  Example 1

The first example contains samples that have either taxon 1 or taxon 2.
```{r Random_Samples1}
tpair1 <- c(selected_pairs$Taxa1[4], selected_pairs$Taxa2[4])
set.seed(2022)
example1 <- cbind(metadata$SAMPLE_ID, metadata$asthma, counts_table[, tpair1]) %>%
  as.data.frame()
colnames(example1)[c(1, 2)] <- c("ID", "asthma")
```

```{r scatterplot1, echo=FALSE, out.width="80%"}
example1$asthma <- as.factor(example1$asthma)
group.colors <- c("#177BB6", "#B63817")
scatter_plot1 <- ggplot(example1, aes_string(x=tpair1[1], y=tpair1[2], color="asthma")) +
    geom_point(position="jitter", size=0.7, alpha=0.7) + 
  scale_colour_manual(values = group.colors) +
    xlab(tpair1[1]) + ylab(tpair1[2])
scatter_plot1
```

The summary of the effect of asthma, the p values and the log likelihood is
```{r}
types <- c("PIRLS", "LA_NM", "GQ_NM", "LA_bobyqa", "GQ_bobyqa")

summarize_glmm1 <- function(number_AGQ, optim_method){
  newmodel <- glmer(cbind(g__Streptobacillus, g__Kingella) ~ asthma + (1|ID), data=example1, family="binomial",
                      nAGQ = number_AGQ, control=lme4::glmerControl(optimizer=optim_method,
                                            optCtrl=list(maxfun=5e5)))
  newmodel_summary <- summary(newmodel)
  c(newmodel_summary$coefficients[2, c(1, 4)], newmodel_summary$AICtab[3])
}

model1_PIRLS <-  summarize_glmm1(0, "Nelder_Mead")
model1_LA_NM <- summarize_glmm1(1, "Nelder_Mead")
model1_GQ_NM <- summarize_glmm1(10, "Nelder_Mead")
model1_LA_bobyqa <- summarize_glmm1(1, "bobyqa")
model1_GQ_bobyqa <- summarize_glmm1(10, 'bobyqa')

output_combined1 <- rbind(model1_PIRLS, model1_LA_NM, model1_GQ_NM,
                           model1_LA_bobyqa, model1_GQ_bobyqa) 
example1_summary <- cbind(types, output_combined1) %>% as.data.frame()
example1_summary[, c(2,3,4)] <- sapply(example1_summary[, c(2,3,4)], as.numeric)
knitr::kable(example1_summary, digits=3)
```



## Example 2

The second example has similar zero-inflated pattern as the first example. 

```{r Random_Samples2, echo=FALSE}
tpair2 <- c(selected_pairs$Taxa1[5], selected_pairs$Taxa2[5])
set.seed(2022)
example2 <- cbind(metadata$SAMPLE_ID, metadata$asthma, counts_table[, tpair2]) %>%
  as.data.frame()
colnames(example2)[c(1, 2)] <- c("ID", "asthma")
```

```{r scatterplot2, echo=FALSE, out.width="80%"}
example2$asthma <- as.factor(example2$asthma)
scatter_plot2 <- ggplot(example2, aes_string(x=tpair2[1], y=tpair2[2], color="asthma")) +
    geom_point(position="jitter", size=0.7, alpha=0.7) + 
  scale_colour_manual(values = group.colors) +
    xlab(tpair2[1]) + ylab(tpair2[2])
scatter_plot2
```

The summary of the effect of $X$, the p values and the log likelihood is
```{r, echo=FALSE}
summarize_glmm2 <- function(number_AGQ, optim_method){
  newmodel <- glmer(cbind(g__Saccharimonadales, g__Moraxella) ~ asthma + (1|ID), data=example2, family="binomial",
                      nAGQ = number_AGQ, control=lme4::glmerControl(optimizer=optim_method,
                                            optCtrl=list(maxfun=2e5)))
  newmodel_summary <- summary(newmodel)
  c(newmodel_summary$coefficients[2, c(1, 4)], newmodel_summary$AICtab[3])
}

model2_PIRLS <-  summarize_glmm2(0, "Nelder_Mead")
model2_LA_NM <- summarize_glmm2(1, "Nelder_Mead")
model2_GQ_NM <- summarize_glmm2(10, "Nelder_Mead")
model2_LA_bobyqa <- summarize_glmm2(1, "bobyqa")
model2_GQ_bobyqa <- summarize_glmm2(10, 'bobyqa')


outcome_combined2 <- rbind(model2_PIRLS, model2_LA_NM, model2_GQ_NM,
                           model2_LA_bobyqa, model2_GQ_bobyqa)

example2_summary <- cbind(types, outcome_combined2) %>% as.data.frame()
colnames(example2_summary) <- c("type", "Effect", "Pval", "Loglik")

example2_summary[, c(2,3,4)] <- sapply(example2_summary[, c(2,3,4)], as.numeric)
knitr::kable(example2_summary, digits=3)
```


# Discussion

1. Both nonlinear optimization algorithm might converge to a wrong local minimum when combined with any approximation method of log likelihood.
2. Nonlinear optimization algorithms might fail to converge, leading to a p value of zero.
3. Using penalized iterative reweighting method combined with Laplace approximation isn't accurate, but doesn't make erroneous conclusions about the significance of asthma effect.


