---
title: "Differential Ratio Analysis on Yatsunenko Data"
author: "Mukai Wang"
date: "3/10/2022"
output:
  html_document:
    df_print: paged
---
```{r, message=FALSE}
library(DiffRatio)
library(biomformat)
library(doParallel)
library(dplyr)
library(igraph)
```

This is a trial of using the `dfr` function in my package to analyze a small subset of Yatsunenko gut microbiome data.

First read in the metadata and the count matrix.
```{r}
folder = '/home/wangmk/UM/Research/MDAWG/Differential_Ratio/data'
metadata = read.csv(file.path(folder, 'metadata.txt'), sep='\t')
OTU = read_biom(file.path(folder, 'study_850', 'BIOM',  '62018', 'all.biom'))
otumat <- as(biom_data(OTU), "matrix")
otumat<-t(otumat)
```

Take a look at the rows and columns of metadata and count matrix.
```{r}
colnames(metadata)
```
```{r}
dim(otumat)
rownames(otumat)[1:5]
colnames(otumat)[1:5]
```

Following [Shulei Wang's approach](https://github.com/lakerwsl/MsRDB-Manuscript-Code/blob/main/MsRDB/Simulation.R), I remove taxa that are only present in fewer than 3% of all the samples.

```{r}
otumat <- otumat[,apply(otumat, 2, function(x){mean(x>0)})>0.03]
otudf <- as.data.frame(otumat)
dim(otudf)
```

Based on the rowids in the count matrix and the metadata, I can identify the country each sample comes from.
```{r}
country <- metadata$country
samplenames <- metadata$sample_name
names(country) <- samplenames

## rename the column names from long DNA sequences to T[ID]
colnames(otudf) <- paste0('T', seq(1, ncol(otudf)))
population <- country[rownames(otumat)]
otudf$origin <- population
table(otudf$origin)
```

In this notebook let's focus on the Malawi and USA populations.

```{r}
otudf_subset <- otudf %>% filter(origin %in% c('Malawi', 'USA'))
otudf_subset$country <- otudf_subset$origin == 'Malawi'
```

We run pairwise differential ratio analysis between taxa T1 to T20.

```{r, cache=TRUE}
tnames <- paste0('T', seq(1, 20))
taxa_pairs <- combn(tnames, 2)

registerDoParallel(cores=4)
duration <- system.time(
  results <- foreach(j=1:ncol(taxa_pairs), .inorder=FALSE) %dopar% {
    model <- dfr(otudf_subset, covar='country', tpair=taxa_pairs[, j])
    coefs <- summary(model)$coefficients
    output <- list(tpair = taxa_pairs[, j],
                   coefficient=coefs[2, 'Estimate'],
                   pval = coefs[2, 'Pr(>|z|)'])
    output
  }
)
duration
```

We adjust P values using Benjamini-Hochberg procedure.

```{r}
pvals <- sapply(results, function(res) res$pval)
adjusted_pvals <- p.adjust(pvals, method='BH')
mean(adjusted_pvals < 0.05)
```

I aggregate the outputs from each model to plot a network indicating differential relative abundance ratio relationships. The plotting code are inspired from [here](https://www.r-graph-gallery.com/249-igraph-network-map-a-color.html).

```{r}
# aggregating results
taxa_pairs <- sapply(results, function(res) res$tpair) %>% t() %>% as.data.frame()
coefficients <- sapply(results, function(res) res$coefficient)
result_df <- cbind(taxa_pairs, coefficients, adjusted_pvals)

# set up the network
links <- result_df %>% filter(adjusted_pvals < 0.05) %>%
  select(V1, V2, coefficients) %>% rename(source = V1, target = V2, importance = coefficients)
nodes <- data.frame(name=paste0('T', seq(1, 20)))
network <- graph_from_data_frame(d=links, directed=TRUE, vertices=nodes)

# plot the network
library(RColorBrewer)
coul  <- brewer.pal(3, "Set1")
colorsetting <- (E(network)$importance > 0) + 1
my_color <- coul[colorsetting]
plot(network,  edge.color=my_color, edge.arrow.size=0.5)
```

Red nodes (e.g. those that point to T19) means that the relative abundance ratio is significantly lower in the Malawi population than in the US population.


Some questions:
1. The current implementation of my package seems slow. How do I improve speed?
2. There are many *.biom files in the downloaded data from QIITA. Which file(s) should I use?

TODO:
1. Rigorously compare theory behind logistic regression and OLS on log ratio.
2. Search for literature for network analysis and summary.
